---
title: "Customer Cohort Retention Analysis"
author: "Victor Gomes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

# Executive Summary

This analysis tracks customer retention patterns by cohort (first purchase month) to identify when customers churn and how retention varies over time.

**Key Objectives:**
- Measure retention rates by cohort and time period
- Identify critical drop-off points in customer lifecycle
- Compare cohort performance over time
- Establish retention benchmarks for the business
- Provide actionable recommendations to improve retention

---

# Introduction to Cohort Analysis

## What is Cohort Analysis?

**Cohort:** A group of customers who made their first purchase in the same month.

**Why Track Cohorts?**
- Understand customer lifecycle and churn patterns
- Identify when customers typically stop purchasing
- Measure effectiveness of retention initiatives
- Compare performance of different customer groups
- Forecast long-term customer value

## Retention Metrics

**Month 0:** First purchase month (100% retention by definition)  
**Month 1:** % of cohort who purchased in 2nd month  
**Month N:** % of cohort still purchasing N months after first purchase

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 12, fig.height = 8, fig.align = 'center')
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(here)
library(scales)
library(knitr)
library(kableExtra)
library(viridis)
```

---

# Data Loading

```{r load-data}
# Load transaction data (customers only)
transactions <- read_csv(here("data", "processed", "retail_customers_only.csv"))

cat("Data loaded:", nrow(transactions), "transactions\n")
cat("Unique customers:", n_distinct(transactions$CustomerID), "\n")
cat("Date range:", as.character(min(as.Date(transactions$InvoiceDate))), "to", 
    as.character(max(as.Date(transactions$InvoiceDate))), "\n")
```

---

# Cohort Assignment

```{r cohort-assignment}
# Assign each customer to cohort based on first purchase
customer_cohorts <- transactions %>%
  filter(!IsReturn) %>%  # Exclude returns
  group_by(CustomerID) %>%
  summarize(
    FirstPurchase = min(InvoiceDate),
    CohortMonth = floor_date(min(InvoiceDate), "month"),
    .groups = "drop"
  ) %>%
  mutate(Cohort = format(CohortMonth, "%Y-%m"))

# Cohort sizes
cohort_sizes <- customer_cohorts %>%
  count(Cohort, name = "CohortSize")

cohort_sizes %>%
  arrange(Cohort) %>%
  kable(caption = "Customer Acquisition by Cohort Month") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\nTotal cohorts:", n_distinct(customer_cohorts$Cohort), "\n")
cat("Average cohort size:", round(mean(cohort_sizes$CohortSize)), "\n")
```

---

# Customer Activity Tracking

```{r activity-tracking}
# Track all purchase months for each customer
customer_activity <- transactions %>%
  filter(!IsReturn) %>%
  mutate(PurchaseMonth = floor_date(InvoiceDate, "month")) %>%
  distinct(CustomerID, PurchaseMonth) %>%
  inner_join(customer_cohorts, by = "CustomerID") %>%
  mutate(
    MonthsSinceCohort = interval(CohortMonth, PurchaseMonth) %/% months(1)
  ) %>%
  filter(MonthsSinceCohort >= 0)  # Sanity check

cat("Total customer-month combinations:", nrow(customer_activity), "\n")
```

---

# Retention Calculation

```{r retention-calculation}
# Calculate retention rates for each cohort and month
retention_data <- customer_activity %>%
  count(Cohort, MonthsSinceCohort, name = "ActiveCustomers") %>%
  left_join(cohort_sizes, by = "Cohort") %>%
  mutate(RetentionRate = round(ActiveCustomers / CohortSize * 100, 1)) %>%
  arrange(Cohort, MonthsSinceCohort)

# Preview
head(retention_data, 15) %>%
  kable(caption = "Sample Retention Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Retention Heatmap (The Iconic Visualization)

```{r retention-heatmap, fig.width=14, fig.height=10}
# Create retention matrix for heatmap
retention_matrix <- retention_data %>%
  select(Cohort, MonthsSinceCohort, RetentionRate) %>%
  pivot_wider(names_from = MonthsSinceCohort, values_from = RetentionRate, 
              names_prefix = "Month_") %>%
  arrange(Cohort)

# Convert back to long format for ggplot
retention_long <- retention_matrix %>%
  pivot_longer(cols = starts_with("Month_"), names_to = "Month", 
               values_to = "Retention") %>%
  mutate(Month = as.numeric(str_remove(Month, "Month_")))

# Create heatmap
ggplot(retention_long, aes(x = Month, y = fct_rev(Cohort), fill = Retention)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(!is.na(Retention), paste0(Retention, "%"), "")), 
            size = 2.5, color = "white", fontface = "bold") +
  scale_fill_viridis(option = "plasma", na.value = "grey90", 
                     limits = c(0, 100), breaks = seq(0, 100, 20)) +
  scale_x_continuous(breaks = seq(0, max(retention_long$Month, na.rm = TRUE), 1)) +
  labs(
    title = "Customer Retention Heatmap by Cohort",
    subtitle = "Percentage of customers still purchasing N months after first purchase",
    x = "Months Since First Purchase",
    y = "Cohort (First Purchase Month)",
    fill = "Retention %",
    caption = "Darker colors = higher retention | Each cell shows % of cohort still active"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.y = element_text(size = 9),
    legend.position = "right",
    panel.grid = element_blank()
  )
```

**Key Patterns to Observe:**

- **Diagonal decline:** Natural retention degradation over time
- **First column:** Always 100% (month of first purchase)
- **Color intensity:** Darker = better retention
- **Horizontal patterns:** Certain cohorts performing consistently better/worse
- **Vertical patterns:** Specific months where all cohorts drop

---

# Retention Curves

```{r retention-curves, fig.width=12, fig.height=7}
# Plot all cohort retention curves
ggplot(retention_data, aes(x = MonthsSinceCohort, y = RetentionRate, 
                           group = Cohort, color = Cohort)) +
  geom_line(alpha = 0.6, linewidth = 1) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  scale_x_continuous(breaks = seq(0, max(retention_data$MonthsSinceCohort), 1)) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Retention Curves by Cohort",
    subtitle = "How customer retention evolves over time for each cohort",
    x = "Months Since First Purchase",
    y = "Retention Rate (%)",
    color = "Cohort"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "right"
  )
```

**Insights:**

- **Convergence:** Do all cohorts eventually stabilize at similar retention?
- **Spread:** Wide variation suggests cohort-specific factors
- **Shape:** Steep vs. gradual decline indicates churn patterns

---

# Average Retention Curve

```{r average-retention, fig.width=12, fig.height=7}
# Calculate average retention by month
avg_retention <- retention_data %>%
  group_by(MonthsSinceCohort) %>%
  summarize(
    AvgRetention = mean(RetentionRate),
    MedianRetention = median(RetentionRate),
    MinRetention = min(RetentionRate),
    MaxRetention = max(RetentionRate),
    .groups = "drop"
  )

# Plot with confidence band
ggplot(avg_retention, aes(x = MonthsSinceCohort)) +
  geom_ribbon(aes(ymin = MinRetention, ymax = MaxRetention), 
              fill = "#1976D2", alpha = 0.2) +
  geom_line(aes(y = AvgRetention), color = "#1976D2", linewidth = 1.5) +
  geom_line(aes(y = MedianRetention), color = "#D32F2F", 
            linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = AvgRetention), color = "#1976D2", size = 3) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  annotate("text", x = max(avg_retention$MonthsSinceCohort) * 0.7, y = 80,
           label = "Blue = Average\nRed = Median\nShaded = Range", 
           hjust = 0, size = 4) +
  labs(
    title = "Average Customer Retention Over Time",
    subtitle = "Cross-cohort average retention with min-max range",
    x = "Months Since First Purchase",
    y = "Retention Rate (%)",
    caption = "Shaded area shows variation across cohorts"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 16))

# Display metrics table
avg_retention %>%
  kable(digits = 1, caption = "Average Retention Metrics by Month",
        col.names = c("Month", "Avg %", "Median %", "Min %", "Max %")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Cohort Size Context

```{r cohort-sizes, fig.width=12, fig.height=6}
# Visualize cohort sizes over time
ggplot(cohort_sizes, aes(x = Cohort, y = CohortSize)) +
  geom_col(fill = "#1976D2", alpha = 0.8) +
  geom_text(aes(label = CohortSize), vjust = -0.5, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Customer Acquisition by Cohort Month",
    subtitle = "Number of new customers per month (context for retention rates)",
    x = "Cohort Month",
    y = "Number of Customers",
    caption = "Larger cohorts provide more reliable retention percentages"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

---

# Key Retention Metrics

```{r key-metrics}
# Calculate critical retention milestones
key_metrics <- avg_retention %>%
  filter(MonthsSinceCohort %in% c(1, 3, 6, 12))

cat("CRITICAL RETENTION BENCHMARKS\n")
cat(rep("=", 60), "\n\n", sep = "")

for(i in 1:nrow(key_metrics)) {
  month <- key_metrics$MonthsSinceCohort[i]
  rate <- key_metrics$AvgRetention[i]
  cat(sprintf("Month %d Retention: %.1f%%\n", month, rate))
}

# Calculate retention half-life
half_life <- avg_retention %>%
  filter(AvgRetention <= 50) %>%
  slice(1)

if(nrow(half_life) > 0) {
  cat(sprintf("\nRetention Half-Life: Month %d (%.1f%% retention)\n", 
              half_life$MonthsSinceCohort, half_life$AvgRetention))
  cat("(Time when 50% of customers have churned)\n")
}

# Calculate plateau retention
plateau <- avg_retention %>%
  filter(MonthsSinceCohort >= 6) %>%
  summarize(PlateauRetention = mean(AvgRetention))

cat(sprintf("\nPlateau Retention (6+ months): %.1f%%\n", plateau$PlateauRetention))
cat("(Long-term steady-state retention rate)\n\n")
```

---

# Cohort Performance Comparison

```{r cohort-comparison}
# Compare best vs worst cohorts
cohort_performance <- retention_data %>%
  filter(MonthsSinceCohort <= 6) %>%  # Focus on first 6 months
  group_by(Cohort) %>%
  summarize(
    AvgRetention = mean(RetentionRate),
    Month1 = RetentionRate[MonthsSinceCohort == 1][1],
    Month3 = RetentionRate[MonthsSinceCohort == 3][1],
    Month6 = RetentionRate[MonthsSinceCohort == 6][1],
    .groups = "drop"
  ) %>%
  left_join(cohort_sizes, by = "Cohort") %>%
  arrange(desc(AvgRetention))

cat("TOP 5 BEST PERFORMING COHORTS (6-Month Avg):\n\n")
cohort_performance %>%
  head(5) %>%
  kable(digits = 1, col.names = c("Cohort", "Avg Retention", "M1", "M3", "M6", "Size")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

cat("\n\nBOTTOM 5 COHORTS:\n\n")
cohort_performance %>%
  tail(5) %>%
  kable(digits = 1, col.names = c("Cohort", "Avg Retention", "M1", "M3", "M6", "Size")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()
```

---

# Critical Drop-Off Analysis

```{r drop-off-analysis}
# Calculate month-over-month retention drop
retention_drops <- avg_retention %>%
  arrange(MonthsSinceCohort) %>%
  mutate(
    RetentionDrop = lag(AvgRetention) - AvgRetention,
    DropPct = round((RetentionDrop / lag(AvgRetention)) * 100, 1)
  ) %>%
  filter(!is.na(RetentionDrop))

cat("RETENTION DROP-OFF BY MONTH\n")
cat(rep("=", 60), "\n\n", sep = "")

retention_drops %>%
  select(MonthsSinceCohort, AvgRetention, RetentionDrop, DropPct) %>%
  arrange(desc(RetentionDrop)) %>%
  head(5) %>%
  kable(digits = 1, 
        col.names = c("Month", "Retention %", "Drop (pp)", "Drop (%)"),
        caption = "Top 5 Months with Steepest Retention Decline") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r drop-off-viz, fig.width=12, fig.height=6}
# Visualize retention drops
ggplot(retention_drops, aes(x = MonthsSinceCohort, y = RetentionDrop)) +
  geom_col(fill = "#D32F2F", alpha = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", RetentionDrop)), 
            vjust = -0.5, size = 3) +
  labs(
    title = "Month-Over-Month Retention Drop",
    subtitle = "How much retention decreases each month (percentage points)",
    x = "Months Since First Purchase",
    y = "Retention Drop (percentage points)",
    caption = "Larger bars indicate critical churn periods"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 16))
```

**Critical intervention windows:** Months with steepest drops are where retention efforts have highest ROI.

---

# Business Insights & Recommendations

```{r insights}
# Calculate key statistics for recommendations
first_month_retention <- avg_retention %>% 
  filter(MonthsSinceCohort == 1) %>% 
  pull(AvgRetention)

churn_first_month <- 100 - first_month_retention

three_month_retention <- avg_retention %>% 
  filter(MonthsSinceCohort == 3) %>% 
  pull(AvgRetention)

best_cohort <- cohort_performance %>% slice(1)
worst_cohort <- cohort_performance %>% slice(n())

cat("\n")
cat(rep("=", 70), "\n")
cat("                    KEY BUSINESS INSIGHTS\n")
cat(rep("=", 70), "\n\n")

cat("1. CRITICAL FIRST MONTH\n")
cat(sprintf("   • %.1f%% of customers make a second purchase\n", first_month_retention))
cat(sprintf("   • %.1f%% churn after first purchase (MAJOR OPPORTUNITY)\n", churn_first_month))
cat("   • Recommendation: Aggressive 2nd purchase incentive program\n\n")

cat("2. RETENTION MILESTONES\n")
cat(sprintf("   • 3-month retention: %.1f%%\n", three_month_retention))
cat(sprintf("   • 6-month retention: %.1f%%\n", 
            avg_retention %>% filter(MonthsSinceCohort == 6) %>% pull(AvgRetention)))
cat("   • Recommendation: Set quarterly retention targets\n\n")

cat("3. COHORT PERFORMANCE VARIATION\n")
cat(sprintf("   • Best cohort (%s): %.1f%% avg retention\n", 
            best_cohort$Cohort, best_cohort$AvgRetention))
cat(sprintf("   • Worst cohort (%s): %.1f%% avg retention\n", 
            worst_cohort$Cohort, worst_cohort$AvgRetention))
cat(sprintf("   • Difference: %.1f percentage points\n", 
            best_cohort$AvgRetention - worst_cohort$AvgRetention))
cat("   • Recommendation: Analyze best cohort acquisition sources\n\n")

cat("4. CRITICAL INTERVENTION WINDOWS\n")
steepest_drop <- retention_drops %>% arrange(desc(RetentionDrop)) %>% slice(1)
cat(sprintf("   • Steepest drop: Month %d (%.1f pp decline)\n", 
            steepest_drop$MonthsSinceCohort, steepest_drop$RetentionDrop))
cat("   • Recommendation: Focus retention campaigns on this period\n\n")

cat("5. LONG-TERM LOYALTY\n")
cat(sprintf("   • Plateau retention: %.1f%%\n", plateau$PlateauRetention))
cat("   • These are \"lifetime\" customers\n")
cat("   • Recommendation: VIP program for customers reaching 6+ months\n\n")

cat(rep("=", 70), "\n\n")
```

---

# Strategic Recommendations

```{r recommendations}
cat("ACTIONABLE RETENTION STRATEGIES\n")
cat(rep("=", 70), "\n\n")

cat("🎯 PRIORITY 1: Improve First-Month Retention\n")
cat(sprintf("   Current: %.1f%% | Target: 55-60%%\n", first_month_retention))
cat("   Actions:\n")
cat("   • Welcome email series (Days 0, 3, 7, 14, 30)\n")
cat("   • 15-20% discount on 2nd purchase within 30 days\n")
cat("   • Product education and usage tips\n")
cat("   • Satisfaction check-in after first purchase\n")
cat(sprintf("   • Expected impact: +10pp retention = £X additional revenue\n\n"))

cat("🎯 PRIORITY 2: Prevent 3-Month Churn\n")
cat(sprintf("   Current: %.1f%% | Target: 35-40%%\n", three_month_retention))
cat("   Actions:\n")
cat("   • Triggered campaign at 60 days since last purchase\n")
cat("   • Personalized product recommendations\n")
cat("   • Early warning system for declining activity\n")
cat("   • Exclusive offers for at-risk customers\n\n")

cat("🎯 PRIORITY 3: Replicate Best Cohort Success\n")
cat(sprintf("   Best cohort: %s (%.1f%% retention)\n", 
            best_cohort$Cohort, best_cohort$AvgRetention))
cat("   Actions:\n")
cat("   • Analyze acquisition source of best cohort\n")
cat("   • Identify product/campaign mix during that period\n")
cat("   • Replicate successful strategies\n")
cat("   • Test on new cohorts and measure\n\n")

cat("🎯 PRIORITY 4: Build Long-Term Loyalty Program\n")
cat(sprintf("   Target: Increase %.1f%% plateau to 30%%+\n", plateau$PlateauRetention))
cat("   Actions:\n")
cat("   • Tiered loyalty program (Bronze/Silver/Gold)\n")
cat("   • Exclusive benefits for 6+ month customers\n")
cat("   • Community building (forum, events)\n")
cat("   • VIP customer service\n\n")

cat(rep("=", 70), "\n\n")
```

---

# Summary & Next Steps

```{r summary}
cat("COHORT ANALYSIS COMPLETE\n")
cat(rep("=", 70), "\n\n")

cat("✅ Analyzed", n_distinct(customer_cohorts$Cohort), "customer cohorts\n")
cat("✅ Tracked retention across", 
    max(retention_data$MonthsSinceCohort), "months\n")
cat(sprintf("✅ Identified %.1f%% first-month retention (critical metric)\n", 
            first_month_retention))
cat(sprintf("✅ Established %.1f%% long-term retention benchmark\n", 
            plateau$PlateauRetention))
cat("✅ Pinpointed critical intervention windows\n\n")

cat("NEXT STEPS:\n")
cat("  1. Implement first-purchase follow-up campaign (immediate)\n")
cat("  2. Set up monthly retention monitoring dashboard\n")
cat("  3. Test retention initiatives on new cohorts\n")
cat("  4. Deep dive into best vs. worst cohort differences\n")
cat("  5. Integrate retention metrics into RFM segmentation\n\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```

---

**Cohort Analysis Complete!**

*This analysis provides the foundation for data-driven retention strategies and customer lifetime value optimization.*