---
title: "Product Affinity Analysis (Market Basket)"
author: "Victor Gomes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

# Executive Summary

This analysis identifies products frequently purchased together to optimize cross-sell strategies, bundle pricing, and product recommendations.

**Objectives:**
- Discover product associations and purchase patterns
- Identify cross-sell and upsell opportunities
- Recommend product bundles
- Optimize product placement and marketing
- Quantify revenue opportunities from recommendations

---

# Introduction to Market Basket Analysis

## What is Market Basket Analysis?

**Market Basket Analysis:** Technique to discover relationships between products by analyzing which items are frequently purchased together.

**Core Metrics:**

**Support:** How often items appear together
```
Support(A,B) = Transactions containing both A and B / Total transactions
```

**Confidence:** Given A was purchased, probability B is also purchased
```
Confidence(A→B) = Transactions with both A and B / Transactions with A
```

**Lift:** How much more likely B is purchased when A is purchased
```
Lift(A→B) = Confidence(A→B) / Support(B)
```
- Lift > 1: Positive association (A increases likelihood of B)
- Lift = 1: No association (independent)
- Lift < 1: Negative association (A decreases likelihood of B)

## Business Applications

- **Cross-sell recommendations:** "Customers who bought X also bought Y"
- **Product bundling:** Create packages of complementary items
- **Store layout:** Place associated products near each other
- **Promotional strategies:** Bundle discount offers
- **Inventory management:** Stock complementary products together

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8, fig.align = 'center')
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(here)
library(scales)
library(knitr)
library(kableExtra)
library(viridis)
library(arules)      # Association rules mining
library(arulesViz)   # Visualization of rules
library(igraph)      # Network graphs
library(ggraph)      # Graph visualization
```

---

# Data Loading

```{r load-data}
# Load transaction data
transactions <- read_csv(here("data", "processed", "retail_customers_only.csv"))

cat("Data loaded:\n")
cat("  Transactions:", nrow(transactions), "\n")
cat("  Unique customers:", n_distinct(transactions$CustomerID), "\n")
cat("  Unique products:", n_distinct(transactions$StockCode), "\n")
cat("  Unique invoices:", n_distinct(transactions$InvoiceNo), "\n")
```

---

# Data Preparation for Market Basket

```{r data-prep}
# Filter to valid sales only (exclude returns, admin items)
basket_data <- transactions %>%
  filter(
    !IsReturn,
    Quantity > 0,
    UnitPrice > 0,
    !is.na(Description),
    Description != ""
  ) %>%
  # Clean product descriptions
  mutate(
    Product = str_trim(str_to_upper(Description)),
    Product = str_replace_all(Product, "[^A-Z0-9 ]", "")  # Remove special chars
  ) %>%
  # Remove admin items
  filter(!StockCode %in% c("POST", "D", "M", "BANK CHARGES", "AMAZONFEE", "DOT", "PADS", "C2", "CRUK"))

cat("\nBasket Data Summary:\n")
cat("  Valid transactions:", nrow(basket_data), "\n")
cat("  Unique products:", n_distinct(basket_data$Product), "\n")
cat("  Unique baskets:", n_distinct(basket_data$InvoiceNo), "\n")
cat("  Average items per basket:", round(nrow(basket_data) / n_distinct(basket_data$InvoiceNo), 2), "\n")
```

---

# Product Popularity Analysis

Before finding associations, understand which products are most popular.

```{r product-popularity}
# Calculate product metrics
product_stats <- basket_data %>%
  group_by(StockCode, Product) %>%
  summarize(
    Times_Purchased = n(),
    Unique_Customers = n_distinct(CustomerID),
    Unique_Baskets = n_distinct(InvoiceNo),
    Total_Quantity = sum(Quantity),
    Total_Revenue = sum(TotalAmount),
    Avg_Price = mean(UnitPrice),
    .groups = "drop"
  ) %>%
  arrange(desc(Unique_Baskets))

# Top 20 products
cat("\nTop 20 Most Popular Products:\n\n")
product_stats %>%
  head(20) %>%
  mutate(
    Total_Revenue = dollar(Total_Revenue, prefix = "£"),
    Avg_Price = dollar(Avg_Price, prefix = "£")
  ) %>%
  select(Product, Unique_Baskets, Unique_Customers, Total_Revenue) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) %>%
  scroll_box(width = "100%", height = "500px")
```

```{r popularity-viz, fig.width=12, fig.height=8}
# Visualize top products
product_stats %>%
  head(25) %>%
  mutate(Product = str_trunc(Product, 40)) %>%
  ggplot(aes(x = Unique_Baskets, y = reorder(Product, Unique_Baskets))) +
  geom_col(fill = "#1976D2", alpha = 0.8) +
  geom_text(aes(label = comma(Unique_Baskets)), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 25 Most Frequently Purchased Products",
    subtitle = "By number of baskets (transactions) containing the product",
    x = "Number of Baskets",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.major.y = element_blank()
  )
```

---

# Basket Size Distribution

```{r basket-size}
# Analyze basket sizes
basket_sizes <- basket_data %>%
  group_by(InvoiceNo) %>%
  summarize(
    Basket_Size = n(),
    Basket_Value = sum(TotalAmount),
    .groups = "drop"
  )

cat("\nBasket Size Statistics:\n")
cat("  Average basket size:", round(mean(basket_sizes$Basket_Size), 2), "items\n")
cat("  Median basket size:", median(basket_sizes$Basket_Size), "items\n")
cat("  Average basket value:", dollar(mean(basket_sizes$Basket_Value), prefix = "£"), "\n")
cat("  Median basket value:", dollar(median(basket_sizes$Basket_Value), prefix = "£"), "\n")
```

```{r basket-size-viz, fig.width=12, fig.height=6}
# Visualize basket size distribution
ggplot(basket_sizes, aes(x = Basket_Size)) +
  geom_histogram(binwidth = 1, fill = "#1976D2", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(Basket_Size)), color = "#D32F2F", 
             linetype = "dashed", linewidth = 1) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, 5)) +
  annotate("text", x = mean(basket_sizes$Basket_Size) + 3, y = Inf, 
           label = paste("Mean:", round(mean(basket_sizes$Basket_Size), 1)), 
           vjust = 2, color = "#D32F2F", fontface = "bold") +
  labs(
    title = "Distribution of Basket Sizes",
    subtitle = "Number of items per transaction",
    x = "Items per Basket",
    y = "Number of Transactions"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

---

# Product Co-Occurrence Matrix

Calculate how often products appear together in baskets.

```{r co-occurrence}
# Focus on top 50 products for manageable analysis
top_products <- product_stats %>%
  head(50) %>%
  pull(Product)

# Create product pairs from same basket
product_pairs <- basket_data %>%
  filter(Product %in% top_products) %>%
  select(InvoiceNo, Product) %>%
  distinct() %>%
  inner_join(., ., by = "InvoiceNo", relationship = "many-to-many") %>%
  filter(Product.x < Product.y) %>%  # Avoid duplicates and self-pairs
  count(Product.x, Product.y, name = "Co_Occurrences") %>%
  arrange(desc(Co_Occurrences))

cat("\nProduct Association Summary:\n")
cat("  Total unique product pairs:", nrow(product_pairs), "\n")
cat("  Top pair co-occurrences:", max(product_pairs$Co_Occurrences), "\n")
```

## Top Product Associations

```{r top-associations}
# Calculate association metrics
total_baskets <- n_distinct(basket_data$InvoiceNo)

product_pair_metrics <- product_pairs %>%
  left_join(product_stats %>% select(Product, P1_Baskets = Unique_Baskets), 
            by = c("Product.x" = "Product")) %>%
  left_join(product_stats %>% select(Product, P2_Baskets = Unique_Baskets), 
            by = c("Product.y" = "Product")) %>%
  mutate(
    # Support: % of baskets containing both
    Support = Co_Occurrences / total_baskets,
    
    # Confidence: Given P1, probability of P2
    Confidence_1_to_2 = Co_Occurrences / P1_Baskets,
    Confidence_2_to_1 = Co_Occurrences / P2_Baskets,
    
    # Lift: How much more likely together than independent
    Lift = (Co_Occurrences / total_baskets) / ((P1_Baskets / total_baskets) * (P2_Baskets / total_baskets))
  ) %>%
  filter(Lift > 1, Co_Occurrences >= 10) %>%  # Meaningful associations only
  arrange(desc(Lift))

cat("\nTop 20 Product Associations (by Lift):\n\n")
product_pair_metrics %>%
  head(20) %>%
  mutate(
    Product.x = str_trunc(Product.x, 30),
    Product.y = str_trunc(Product.y, 30),
    Support = percent(Support, accuracy = 0.01),
    Confidence_1_to_2 = percent(Confidence_1_to_2, accuracy = 0.1),
    Lift = round(Lift, 2)
  ) %>%
  select(Product.x, Product.y, Co_Occurrences, Support, Confidence_1_to_2, Lift) %>%
  kable(col.names = c("Product A", "Product B", "Co-occur", "Support", "Confidence", "Lift")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) %>%
  scroll_box(width = "100%", height = "500px")
```

**Interpretation:**
- **Lift > 3:** Strong association (buy together 3x more than expected)
- **Confidence > 50%:** If A purchased, >50% chance B is also purchased
- **Support:** How common the combination is overall

---

# Association Rules Mining

Use Apriori algorithm for systematic rule discovery.

```{r apriori-prep, attr.output=FALSE}
# Prepare transaction data in format required by arules
basket_format <- basket_data %>%
  select(InvoiceNo, Product) %>%
  distinct() %>%
  group_by(InvoiceNo) %>%
  summarize(Products = list(Product), .groups = "drop")

# Convert to transactions format
transactions_list <- basket_format$Products
names(transactions_list) <- basket_format$InvoiceNo

# Create transactions object
trans <- as(transactions_list, "transactions")

cat("\nTransaction Data Summary:\n")
cat("  Total transactions:", length(trans), "\n")
cat("  Total unique items:", length(itemLabels(trans)), "\n")
cat("  Density:", round(arules::size(trans) / (length(trans) * length(itemLabels(trans))) * 100, 2), "%\n")
```

```{r apriori-rules}
# Mine association rules
# Parameters: min support 0.5% (appears in 0.5% of baskets), min confidence 20%
rules <- apriori(trans, 
                 parameter = list(supp = 0.005, conf = 0.20, minlen = 2, maxlen = 3),
                 control = list(verbose = FALSE))

# Sort by lift
rules <- sort(rules, by = "lift", decreasing = TRUE)

cat("\nAssociation Rules Discovered:", length(rules), "\n\n")

# Inspect top rules
cat("Top 20 Association Rules (by Lift):\n\n")
top_rules <- head(rules, 20)
rules_df <- as(top_rules, "data.frame")

rules_df %>%
  mutate(
    rules = str_replace_all(rules, "=>", " → "),
    support = percent(support, accuracy = 0.01),
    confidence = percent(confidence, accuracy = 0.1),
    lift = round(lift, 2)
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) %>%
  scroll_box(width = "100%", height = "500px")
```

---

# Recommendation Opportunities

Identify specific cross-sell opportunities.

```{r recommendations}
# Focus on high-confidence rules for recommendations
strong_rules <- subset(rules, confidence > 0.30 & lift > 2)
strong_rules <- sort(strong_rules, by = "confidence", decreasing = TRUE)

cat("\nHigh-Confidence Recommendations (>30% confidence, Lift >2):\n")
cat("  Total strong rules:", length(strong_rules), "\n\n")

if(length(strong_rules) > 0) {
  recommendations_df <- as(head(strong_rules, 15), "data.frame")
  
  recommendations_df %>%
    mutate(
      rules = str_replace_all(rules, "=>", " → "),
      support = percent(support, accuracy = 0.01),
      confidence = percent(confidence, accuracy = 0.1),
      lift = round(lift, 2)
    ) %>%
    kable(caption = "Top 15 Product Recommendation Rules") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) %>%
    scroll_box(width = "100%")
}
```

---

# Product Network Visualization

Visualize product relationships as a network graph.

```{r network-viz, fig.width=14, fig.height=10}
# Create network from top associations
network_data <- product_pair_metrics %>%
  filter(Lift >= 3, Co_Occurrences >= 20) %>%  # Strong associations only
  head(50) %>%
  mutate(
    Product.x = str_trunc(Product.x, 25),
    Product.y = str_trunc(Product.y, 25)
  )

if(nrow(network_data) > 0) {
  # Create igraph object
  graph <- graph_from_data_frame(
    network_data %>% select(Product.x, Product.y, Lift),
    directed = FALSE
  )
  
  # Calculate node importance (degree)
  V(graph)$degree <- degree(graph)
  
  # Plot network
  ggraph(graph, layout = "fr") +
    geom_edge_link(aes(width = Lift, alpha = Lift), color = "#1976D2") +
    geom_node_point(aes(size = degree), color = "#D32F2F", alpha = 0.7) +
    geom_node_text(aes(label = name), size = 2.5, repel = TRUE, 
                   max.overlaps = 20, fontface = "bold") +
    scale_edge_width(range = c(0.5, 3)) +
    scale_edge_alpha(range = c(0.3, 0.8)) +
    scale_size(range = c(3, 10)) +
    labs(
      title = "Product Affinity Network",
      subtitle = "Products connected by strong associations (Lift ≥ 3)",
      caption = "Node size = centrality | Edge width = association strength"
    ) +
    theme_graph() +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12),
      legend.position = "none"
    )
} else {
  cat("Not enough strong associations for network visualization\n")
}
```

---

# Bundle Recommendations

Suggest product bundles based on associations.

```{r bundle-recommendations}
# Identify 3-item bundles with strong associations
bundle_candidates <- product_pair_metrics %>%
  filter(Lift >= 2.5, Confidence_1_to_2 >= 0.25) %>%
  arrange(desc(Lift)) %>%
  head(30)

cat("\n📦 RECOMMENDED PRODUCT BUNDLES\n")
cat(rep("=", 70), "\n\n")

# Display top bundles
for(i in 1:min(10, nrow(bundle_candidates))) {
  p1 <- str_trunc(bundle_candidates$Product.x[i], 40)
  p2 <- str_trunc(bundle_candidates$Product.y[i], 40)
  lift <- round(bundle_candidates$Lift[i], 2)
  conf <- round(bundle_candidates$Confidence_1_to_2[i] * 100, 1)
  
  cat(sprintf("Bundle %d:\n", i))
  cat(sprintf("  • %s\n", p1))
  cat(sprintf("  • %s\n", p2))
  cat(sprintf("  Association: %.1f%% of %s buyers also buy second item (Lift: %.2f)\n\n", 
              conf, p1, lift))
}
```

---

# Revenue Opportunity Analysis

Quantify potential revenue from product recommendations.

```{r revenue-opportunity}
# Calculate revenue impact of recommendations
cat("\nREVENUE OPPORTUNITY FROM PRODUCT RECOMMENDATIONS\n")
cat(rep("=", 70), "\n\n")

# Analyze baskets that could benefit from recommendations
single_item_baskets <- basket_sizes %>%
  filter(Basket_Size == 1) %>%
  nrow()

small_baskets <- basket_sizes %>%
  filter(Basket_Size <= 3) %>%
  nrow()

cat("Current State:\n")
cat(sprintf("  • Single-item baskets: %s (%.1f%%)\n", 
            comma(single_item_baskets),
            single_item_baskets / nrow(basket_sizes) * 100))
cat(sprintf("  • Small baskets (≤3 items): %s (%.1f%%)\n",
            comma(small_baskets),
            small_baskets / nrow(basket_sizes) * 100))
cat(sprintf("  • Average basket value: %s\n\n", 
            dollar(mean(basket_sizes$Basket_Value), prefix = "£")))

# Conservative estimates
cat("Revenue Opportunity (Conservative Estimates):\n\n")

# Scenario 1: Convert 10% of single-item to 2-item baskets
avg_item_value <- mean(basket_data$TotalAmount)
scenario1_revenue <- single_item_baskets * 0.10 * avg_item_value

cat("Scenario 1: Cross-sell to single-item baskets\n")
cat(sprintf("  • Target: 10%% conversion rate\n"))
cat(sprintf("  • Additional items: %s\n", comma(round(single_item_baskets * 0.10))))
cat(sprintf("  • Estimated revenue: %s\n\n", dollar(scenario1_revenue, prefix = "£")))

# Scenario 2: Increase average basket size by 0.5 items
scenario2_revenue <- nrow(basket_sizes) * 0.5 * avg_item_value * 0.15  # 15% take rate

cat("Scenario 2: Increase average basket size\n")
cat(sprintf("  • Target: +0.5 items per basket average\n"))
cat(sprintf("  • Take rate: 15%% of customers\n"))
cat(sprintf("  • Estimated revenue: %s\n\n", dollar(scenario2_revenue, prefix = "£")))

# Scenario 3: Bundle promotions
top_bundle_volume <- sum(bundle_candidates$Co_Occurrences[1:5])
scenario3_revenue <- top_bundle_volume * 0.20 * avg_item_value  # 20% incremental

cat("Scenario 3: Promote top 5 product bundles\n")
cat(sprintf("  • Bundle discount: 10%% off\n"))
cat(sprintf("  • Expected incremental items: 20%% increase\n"))
cat(sprintf("  • Estimated revenue: %s\n\n", dollar(scenario3_revenue, prefix = "£")))

total_opportunity <- scenario1_revenue + scenario2_revenue + scenario3_revenue
cat(sprintf("💰 TOTAL ANNUAL OPPORTUNITY: %s\n\n", 
            dollar(total_opportunity, prefix = "£")))

cat(rep("=", 70), "\n\n")
```

---

# Strategic Recommendations

```{r strategic-recommendations}
cat("\nACTIONABLE RECOMMENDATIONS\n")
cat(rep("=", 70), "\n\n")

cat("🎯 PRIORITY 1: Implement Recommendation Engine\n")
cat("   Actions:\n")
cat("   • Add \"Customers also bought\" on product pages\n")
cat("   • Show top 3-5 related products based on association rules\n")
cat("   • Display during checkout process\n")
cat("   • A/B test placement and messaging\n")
cat(sprintf("   • Expected impact: %s annual revenue\n\n", 
            dollar(scenario1_revenue + scenario2_revenue, prefix = "£")))

cat("🎯 PRIORITY 2: Create Product Bundles\n")
cat("   Actions:\n")
cat("   • Launch top 10 product bundles with 10% discount\n")
cat("   • Prominent placement on homepage and category pages\n")
cat("   • Email campaign to existing customers\n")
cat("   • Track bundle performance vs individual sales\n")
cat(sprintf("   • Expected impact: %s annual revenue\n\n",
            dollar(scenario3_revenue, prefix = "£")))

cat("🎯 PRIORITY 3: Optimize Product Placement\n")
cat("   Actions:\n")
cat("   • Group associated products in same categories\n")
cat("   • Adjacent placement in catalogs/emails\n")
cat("   • \"Complete the set\" messaging\n")
cat("   • Visual merchandising based on associations\n\n")

cat("🎯 PRIORITY 4: Personalized Email Campaigns\n")
cat("   Actions:\n")
cat("   • Send targeted cross-sell emails based on past purchases\n")
cat("   • \"You might also like\" recommendations\n")
cat("   • Time recommendations based on purchase cycles\n")
cat("   • Segment by customer value (CLV tiers)\n\n")

cat("🎯 PRIORITY 5: Monitor and Optimize\n")
cat("   Actions:\n")
cat("   • Track recommendation click-through rates\n")
cat("   • Measure conversion rates by recommendation type\n")
cat("   • Update association rules monthly\n")
cat("   • A/B test different recommendation strategies\n\n")

cat(rep("=", 70), "\n\n")
```

---

# Export Recommendations

```{r export-recommendations}
# Export recommendation rules for implementation
if(length(strong_rules) > 0) {
  recommendations_export <- as(strong_rules, "data.frame") %>%
    separate(rules, into = c("Antecedent", "Consequent"), sep = " => ") %>%
    mutate(
      Antecedent = str_remove_all(Antecedent, "[{}]"),
      Consequent = str_remove_all(Consequent, "[{}]")
    ) %>%
    arrange(desc(confidence))
  
  export_path <- here("data", "processed", "product_recommendations.csv")
  write_csv(recommendations_export, export_path)
  
  cat("✓ Product recommendations exported to:", export_path, "\n")
  cat("  Rules exported:", nrow(recommendations_export), "\n\n")
  
  cat("Ready for:\n")
  cat("  • E-commerce platform integration\n")
  cat("  • Recommendation engine\n")
  cat("  • Marketing campaign targeting\n")
  cat("  • Merchandising decisions\n")
}
```

---

# Summary & Key Takeaways

```{r summary}
cat("\n")
cat(rep("=", 70), "\n")
cat("            PRODUCT AFFINITY ANALYSIS COMPLETE\n")
cat(rep("=", 70), "\n\n")

cat("✅ ACHIEVEMENTS:\n\n")
cat("  1. Analyzed", n_distinct(basket_data$InvoiceNo), "baskets\n")
cat("  2. Discovered", length(rules), "association rules\n")
cat("  3. Identified", length(strong_rules), "high-confidence recommendations\n")
cat("  4. Created product network showing", 
    nrow(network_data), "strong associations\n")
cat("  5. Quantified revenue opportunity:", 
    dollar(total_opportunity, prefix = "£"), "\n\n")

cat("🎯 KEY INSIGHTS:\n\n")
cat(sprintf("  • Average basket size: %.1f items\n", mean(basket_sizes$Basket_Size)))
cat(sprintf("  • Single-item baskets: %.1f%% (cross-sell opportunity)\n",
            single_item_baskets / nrow(basket_sizes) * 100))
cat(sprintf("  • Top association lift: %.1fx\n", max(product_pair_metrics$Lift)))
cat(sprintf("  • Revenue opportunity: %s annually\n", 
            dollar(total_opportunity, prefix = "£")))

cat("\n💡 NEXT STEPS:\n\n")
cat("  1. Implement recommendation engine on website\n")
cat("  2. Launch top 10 product bundles\n")
cat("  3. A/B test recommendation placements\n")
cat("  4. Set up monthly rule updates\n")
cat("  5. Track recommendation performance metrics\n\n")

cat(rep("=", 70), "\n\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```

---

**✅ Product Affinity Analysis Complete!**

*This analysis enables data-driven cross-sell strategies and product bundling.*