---
title: "Customer Churn Prediction Model"
author: "Victor Gomes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

# Executive Summary

This analysis builds a predictive model to identify customers at risk of churning, enabling proactive retention efforts.

**Objectives:**
- Define and measure customer churn
- Build predictive model using behavioral features
- Identify high-risk customers for intervention
- Quantify revenue at risk from churn
- Provide targeted retention recommendations

---

# Introduction to Churn Prediction

## What is Customer Churn?

**Churn:** When a customer stops purchasing from the business.

**Churn Definition (for this analysis):**
- Customer has not purchased in 180+ days after prediction point
- Based on temporal validation approach

## Why Predict Churn?

**Proactive Retention:**
- Identify at-risk customers before they leave
- Intervene with targeted campaigns
- More cost-effective than re-acquisition

**Business Impact:**
- Reduce customer attrition by 15-25%
- Increase customer lifetime value
- Optimize retention budget allocation
- Improve forecasting accuracy

## Predictive Approach

**Model:** Logistic Regression (interpretable, proven effective)

**Features:** Behavioral indicators from historical data only
- Recency (days since last purchase at prediction point)
- Frequency (purchase count)
- Monetary (total spend)
- Trends (increasing/decreasing activity)
- Engagement patterns

**Temporal Validation:** Features calculated from past data, churn measured in future period

---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8, fig.align = 'center')
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(here)
library(scales)
library(knitr)
library(kableExtra)
library(viridis)
library(caret)       # Machine learning
library(pROC)        # ROC curves
library(ROCR)        # Additional ROC functions
```

---

# Data Loading
```{r load-data}
# Load transaction data
transactions <- read_csv(here("data", "processed", "retail_customers_only.csv"))

cat("Data loaded:\n")
cat("  Transactions:", nrow(transactions), "\n")
cat("  Unique Customers:", n_distinct(transactions$CustomerID), "\n")
```

---

# Define Temporal Split
```{r temporal-split}
# Define prediction point (90 days before dataset end)
analysis_date <- max(transactions$InvoiceDate)
prediction_point <- analysis_date - days(90)
churn_window <- 180  # Days of inactivity = churn

cat("\nTemporal Split Configuration:\n")
cat("  Analysis Date:", as.character(analysis_date), "\n")
cat("  Prediction Point:", as.character(prediction_point), "\n")
cat("  Churn Window:", churn_window, "days\n\n")

# Split data into historical and observation periods
historical_transactions <- transactions %>%
  filter(InvoiceDate <= prediction_point)

observation_transactions <- transactions %>%
  filter(InvoiceDate > prediction_point & InvoiceDate <= prediction_point + days(churn_window))

cat("Data Split:\n")
cat("  Historical transactions:", nrow(historical_transactions), "\n")
cat("  Observation transactions:", nrow(observation_transactions), "\n")
```

---

# Feature Engineering

Create predictive features from historical data only (no data leakage).
```{r feature-engineering}
cat("\nCalculating customer features from historical data...\n")

# Basic customer metrics from historical period
customer_features <- historical_transactions %>%
  filter(!IsReturn) %>%
  group_by(CustomerID) %>%
  summarize(
    # Transaction metrics
    Hist_TotalTransactions = n_distinct(InvoiceNo),
    Hist_TotalSpent = sum(TotalAmount),
    Hist_TotalItems = sum(Quantity),
    Hist_AOV = mean(TotalAmount),
    
    # Temporal metrics
    Hist_FirstPurchase = min(InvoiceDate),
    Hist_LastPurchase = max(InvoiceDate),
    Hist_CustomerLifetimeDays = as.numeric(difftime(max(InvoiceDate), min(InvoiceDate), units = "days")),
    
    # Frequency metrics
    Hist_AvgDaysBetweenPurchases = as.numeric(difftime(max(InvoiceDate), min(InvoiceDate), units = "days")) / 
                                   pmax(n_distinct(InvoiceNo) - 1, 1),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Recency at prediction point
    Hist_RecencyDays = as.numeric(difftime(prediction_point, Hist_LastPurchase, units = "days")),
    
    # Purchase frequency
    Hist_PurchaseFrequency = Hist_TotalTransactions / pmax(Hist_CustomerLifetimeDays / 30, 1)
  )

# Calculate temporal trends
customer_trends <- historical_transactions %>%
  filter(!IsReturn) %>%
  mutate(PurchaseMonth = floor_date(InvoiceDate, "month")) %>%
  group_by(CustomerID, PurchaseMonth) %>%
  summarize(
    Monthly_Transactions = n_distinct(InvoiceNo),
    Monthly_Spend = sum(TotalAmount),
    .groups = "drop"
  ) %>%
  group_by(CustomerID) %>%
  summarize(
    # Activity consistency
    Months_Active = n(),
    Avg_Monthly_Spend = mean(Monthly_Spend),
    Std_Monthly_Spend = sd(Monthly_Spend, na.rm = TRUE),
    
    # Recent vs historical comparison (last 3 months vs earlier)
    Recent_3Mo_Spend = sum(Monthly_Spend[PurchaseMonth >= prediction_point - months(3)]),
    Historical_Spend = sum(Monthly_Spend[PurchaseMonth < prediction_point - months(3)]),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Spending trend
    Spend_Trend = ifelse(Historical_Spend > 0, 
                         (Recent_3Mo_Spend / Historical_Spend) - 1, 0),
    Spend_Trend = pmin(pmax(Spend_Trend, -1), 3),
    
    # Spending volatility
    Spend_Volatility = ifelse(!is.na(Std_Monthly_Spend), Std_Monthly_Spend, 0)
  )

# Combine all features
model_data <- customer_features %>%
  left_join(customer_trends, by = "CustomerID") %>%
  mutate(
    # Handle NAs
    across(c(Months_Active, Avg_Monthly_Spend, Spend_Trend, Spend_Volatility), ~replace_na(., 0)),
    
    # Normalized features
    Recency_Normalized = log1p(Hist_RecencyDays),
    Frequency_Normalized = log1p(Hist_TotalTransactions),
    Monetary_Normalized = log1p(Hist_TotalSpent),
    AOV_Normalized = log1p(Hist_AOV),
    
    # Interaction features
    Recency_Frequency_Ratio = Hist_RecencyDays / pmax(Hist_AvgDaysBetweenPurchases, 1),
    Spend_per_Transaction = Hist_TotalSpent / pmax(Hist_TotalTransactions, 1)
  )

cat("\nFeature Engineering Complete\n")
cat("  Total features:", ncol(model_data), "\n")
cat("  Customers:", nrow(model_data), "\n")
```

---

# Define Churn Labels
```{r define-churn}
cat("Defining churn labels from observation period...\n")

# Identify customers who made purchases in observation period
active_customers <- observation_transactions %>%
  filter(!IsReturn) %>%
  group_by(CustomerID) %>%
  summarize(Had_Purchase = TRUE, .groups = "drop")

# Add churn labels
model_data <- model_data %>%
  left_join(active_customers, by = "CustomerID") %>%
  mutate(
    Churned = is.na(Had_Purchase),
    Churn_Label = ifelse(Churned, "Churned", "Active"),
    Churn_Binary = as.integer(Churned)
  ) %>%
  select(-Had_Purchase, -Hist_FirstPurchase, -Hist_LastPurchase) %>%
  filter(!is.na(Hist_TotalTransactions), Hist_TotalTransactions > 0)

# Churn summary
churn_summary <- model_data %>%
  count(Churn_Label) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))

cat("\nChurn Distribution:\n")
churn_summary %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\nChurn Rate:", churn_summary$Percentage[churn_summary$Churn_Label == "Churned"], "%\n")
```

---

# Exploratory Analysis: Churned vs Active
```{r churn-comparison}
# Compare key metrics between churned and active
comparison <- model_data %>%
  group_by(Churn_Label) %>%
  summarize(
    Customers = n(),
    Avg_Recency = round(mean(Hist_RecencyDays), 1),
    Avg_Frequency = round(mean(Hist_TotalTransactions), 1),
    Avg_Monetary = round(mean(Hist_TotalSpent), 2),
    Avg_AOV = round(mean(Hist_AOV), 2),
    Avg_Lifetime_Days = round(mean(Hist_CustomerLifetimeDays), 0),
    .groups = "drop"
  )

comparison %>%
  mutate(
    Avg_Monetary = dollar(Avg_Monetary, prefix = "£"),
    Avg_AOV = dollar(Avg_AOV, prefix = "£")
  ) %>%
  kable(caption = "Churned vs Active Customer Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r churn-viz, fig.width=12, fig.height=6}
# Visualize key differences with log transformation
behavioral_comparison <- model_data %>%
  select(CustomerID, Churn_Label, Hist_RecencyDays, Hist_TotalSpent, Hist_TotalTransactions) %>%
  mutate(
    Status = Churn_Label,
    TotalSpent_Log = log10(Hist_TotalSpent + 1),
    TotalTransactions_Log = log10(Hist_TotalTransactions + 1)
  ) %>%
  select(Status, Hist_RecencyDays, TotalSpent_Log, TotalTransactions_Log) %>%
  pivot_longer(
    cols = c(Hist_RecencyDays, TotalSpent_Log, TotalTransactions_Log),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = case_when(
      Metric == "Hist_RecencyDays" ~ "Recency (Days Since Last Purchase)",
      Metric == "TotalSpent_Log" ~ "Total Spent (log10)",
      Metric == "TotalTransactions_Log" ~ "Total Transactions (log10)"
    ),
    Metric = factor(Metric, levels = c("Recency (Days Since Last Purchase)", 
                                       "Total Spent (log10)", 
                                       "Total Transactions (log10)"))
  )

ggplot(behavioral_comparison, aes(x = Status, y = Value, fill = Status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  coord_cartesian(ylim = c(0, NA)) +
  facet_wrap(~Metric, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Active" = "#2E7D32", "Churned" = "#D32F2F")) +
  labs(
    title = "Behavioral Differences: Active vs Churned Customers",
    subtitle = "Features calculated from historical period only (no data leakage)",
    x = NULL,
    y = "Value",
    fill = "Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

---

# Train/Test Split
```{r train-test-split}
set.seed(123)  # Reproducibility

# Create training and test sets (70/30 split)
train_index <- createDataPartition(model_data$Churned, p = 0.7, list = FALSE)

train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

cat("\nData Split:\n")
cat("  Training set:", nrow(train_data), "customers\n")
cat("  Test set:", nrow(test_data), "customers\n")
cat("  Training churn rate:", 
    round(sum(train_data$Churned) / nrow(train_data) * 100, 1), "%\n")
cat("  Test churn rate:", 
    round(sum(test_data$Churned) / nrow(test_data) * 100, 1), "%\n")
```

---

# Logistic Regression Model
```{r logistic-model}
# Build logistic regression model with historical features only
model_formula <- as.formula(
  Churned ~ Recency_Normalized + Frequency_Normalized + Monetary_Normalized +
    AOV_Normalized + Spend_Trend + Spend_Volatility + Months_Active +
    Recency_Frequency_Ratio + Spend_per_Transaction
)

logistic_model <- glm(model_formula, 
                      data = train_data, 
                      family = binomial(link = "logit"))

# Model summary
cat("\nLogistic Regression Model Summary:\n\n")
summary(logistic_model)

# Feature importance
feature_importance <- data.frame(
  Feature = names(coef(logistic_model))[-1],
  Coefficient = coef(logistic_model)[-1],
  Abs_Coefficient = abs(coef(logistic_model)[-1])
) %>%
  arrange(desc(Abs_Coefficient))

cat("\n\nFeature Importance (by coefficient magnitude):\n\n")
feature_importance %>%
  mutate(
    Coefficient = round(Coefficient, 4),
    Abs_Coefficient = round(Abs_Coefficient, 4)
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r feature-importance-viz, fig.width=12, fig.height=6}
# Visualize feature importance
ggplot(feature_importance, aes(x = reorder(Feature, Abs_Coefficient), y = Coefficient)) +
  geom_col(aes(fill = Coefficient > 0), alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#D32F2F", "FALSE" = "#2E7D32"),
                    labels = c("Increases Churn Risk", "Decreases Churn Risk")) +
  labs(
    title = "Feature Importance in Churn Prediction",
    subtitle = "Logistic regression coefficients",
    x = "Feature",
    y = "Coefficient",
    fill = "Effect"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

---

# Model Predictions
```{r predictions}
# Predict on test set
test_data$Churn_Probability <- predict(logistic_model, newdata = test_data, type = "response")
test_data$Churn_Prediction <- ifelse(test_data$Churn_Probability > 0.5, TRUE, FALSE)

# Predict on full dataset for business use
model_data$Churn_Probability <- predict(logistic_model, newdata = model_data, type = "response")
model_data$Churn_Prediction <- ifelse(model_data$Churn_Probability > 0.5, TRUE, FALSE)
model_data$Risk_Category <- cut(model_data$Churn_Probability, 
                                 breaks = c(0, 0.25, 0.5, 0.75, 1),
                                 labels = c("Low Risk", "Medium Risk", "High Risk", "Critical Risk"),
                                 include.lowest = TRUE)

cat("\nPredictions Complete\n")
cat("  Average churn probability:", 
    round(mean(model_data$Churn_Probability) * 100, 1), "%\n")
```

---

# Model Evaluation
```{r confusion-matrix}
# Confusion matrix
conf_matrix <- confusionMatrix(
  factor(test_data$Churn_Prediction, levels = c(FALSE, TRUE)),
  factor(test_data$Churned, levels = c(FALSE, TRUE)),
  positive = "TRUE"
)

cat("\nConfusion Matrix:\n\n")
print(conf_matrix$table)

cat("\n\nModel Performance Metrics:\n")
cat("  Accuracy:", round(conf_matrix$overall['Accuracy'] * 100, 1), "%\n")
cat("  Sensitivity (Recall):", round(conf_matrix$byClass['Sensitivity'] * 100, 1), "%\n")
cat("  Specificity:", round(conf_matrix$byClass['Specificity'] * 100, 1), "%\n")
cat("  Precision:", round(conf_matrix$byClass['Precision'] * 100, 1), "%\n")
cat("  F1 Score:", round(conf_matrix$byClass['F1'] * 100, 1), "%\n")
```

## ROC Curve
```{r roc-curve, fig.width=10, fig.height=8}
# ROC curve
roc_obj <- roc(test_data$Churned, test_data$Churn_Probability)
auc_value <- auc(roc_obj)

# Plot ROC
plot(roc_obj, 
     main = paste0("ROC Curve (AUC = ", round(auc_value, 3), ")"),
     col = "#1976D2", 
     lwd = 2,
     print.auc = TRUE,
     print.auc.y = 0.4)
abline(a = 0, b = 1, lty = 2, col = "gray")

cat("\nAUC (Area Under Curve):", round(auc_value, 3), "\n")
cat("Interpretation:", ifelse(auc_value > 0.8, "Excellent", 
                              ifelse(auc_value > 0.7, "Good", "Fair")), "\n")
```

---

# Risk Segmentation
```{r risk-segments}
# Analyze risk categories
risk_summary <- model_data %>%
  group_by(Risk_Category) %>%
  summarize(
    Customers = n(),
    Pct_Customers = round(n() / nrow(model_data) * 100, 1),
    Avg_Churn_Prob = round(mean(Churn_Probability) * 100, 1),
    Actual_Churned = sum(Churned),
    Actual_Churn_Rate = round(sum(Churned) / n() * 100, 1),
    Total_Revenue_at_Risk = sum(Hist_TotalSpent),
    Avg_CLV = mean(Hist_TotalSpent),
    .groups = "drop"
  ) %>%
  arrange(desc(Avg_Churn_Prob))

risk_summary %>%
  mutate(
    Total_Revenue_at_Risk = dollar(Total_Revenue_at_Risk, prefix = "£"),
    Avg_CLV = dollar(Avg_CLV, prefix = "£")
  ) %>%
  kable(caption = "Churn Risk Segmentation") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r risk-viz, fig.width=12, fig.height=7}
# Visualize risk distribution
ggplot(model_data, aes(x = Churn_Probability, fill = Risk_Category)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 1) +
  scale_fill_manual(values = c("Low Risk" = "#2E7D32", "Medium Risk" = "#F57C00",
                                "High Risk" = "#D32F2F", "Critical Risk" = "#7B1FA2")) +
  labs(
    title = "Distribution of Churn Probability Scores",
    subtitle = "Vertical line shows 50% threshold for churn prediction",
    x = "Churn Probability",
    y = "Number of Customers",
    fill = "Risk Category"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

---

# High-Risk Customer Analysis
```{r high-risk-analysis}
# Focus on high and critical risk customers
high_risk <- model_data %>%
  filter(Risk_Category %in% c("High Risk", "Critical Risk"))

cat("\nHIGH-RISK CUSTOMER ANALYSIS\n")
cat(rep("=", 70), "\n\n")

cat("High/Critical Risk Customers:", nrow(high_risk), "\n")
cat("Percentage of base:", round(nrow(high_risk) / nrow(model_data) * 100, 1), "%\n")
cat("Revenue at risk:", dollar(sum(high_risk$Hist_TotalSpent), prefix = "£"), "\n")
cat("Average customer value:", dollar(mean(high_risk$Hist_TotalSpent), prefix = "£"), "\n\n")

# Characteristics of high-risk customers
cat("Behavioral Characteristics:\n")
cat("  Avg days since last purchase:", round(mean(high_risk$Hist_RecencyDays), 1), "\n")
cat("  Avg lifetime purchases:", round(mean(high_risk$Hist_TotalTransactions), 1), "\n")
cat("  Avg total spent:", dollar(mean(high_risk$Hist_TotalSpent), prefix = "£"), "\n")
cat("  Avg months active:", round(mean(high_risk$Months_Active, na.rm = TRUE), 1), "\n")
```

## Top 20 Highest Risk Customers
```{r top-risk-customers}
# Identify top risk customers for immediate action
top_risk <- model_data %>%
  arrange(desc(Churn_Probability)) %>%
  head(20) %>%
  select(CustomerID, Churn_Probability, Risk_Category, Hist_RecencyDays,
         Hist_TotalTransactions, Hist_TotalSpent)

top_risk %>%
  mutate(
    Churn_Probability = percent(Churn_Probability, accuracy = 0.1),
    Hist_TotalSpent = dollar(Hist_TotalSpent, prefix = "£")
  ) %>%
  kable(caption = "Top 20 Customers at Highest Churn Risk") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) %>%
  scroll_box(width = "100%")
```

---

# Business Impact & ROI
```{r business-impact}
cat("\n")
cat(rep("=", 70), "\n")
cat("                    BUSINESS IMPACT ANALYSIS\n")
cat(rep("=", 70), "\n\n")

# Calculate retention campaign scenarios
total_high_risk <- sum(model_data$Risk_Category %in% c("High Risk", "Critical Risk"))
revenue_at_risk <- sum((model_data %>% 
  filter(Risk_Category %in% c("High Risk", "Critical Risk")))$Hist_TotalSpent)

cat("CHURN RISK ASSESSMENT:\n\n")
cat(sprintf("  • High/Critical risk customers: %s\n", comma(total_high_risk)))
cat(sprintf("  • Total revenue at risk: %s\n", dollar(revenue_at_risk, prefix = "£")))
cat(sprintf("  • Average value per at-risk customer: %s\n\n",
            dollar(revenue_at_risk / total_high_risk, prefix = "£")))

# Retention scenarios
cat("RETENTION CAMPAIGN SCENARIOS:\n\n")

# Scenario 1: Save 30% of high-risk customers
save_rate_1 <- 0.30
saved_revenue_1 <- revenue_at_risk * save_rate_1
campaign_cost_1 <- total_high_risk * 20  # £20 per customer

cat("Scenario 1: Conservative (30% save rate)\n")
cat(sprintf("  • Customers saved: %s\n", comma(round(total_high_risk * save_rate_1))))
cat(sprintf("  • Revenue retained: %s\n", dollar(saved_revenue_1, prefix = "£")))
cat(sprintf("  • Campaign cost (£20/customer): %s\n", dollar(campaign_cost_1, prefix = "£")))
cat(sprintf("  • Net benefit: %s\n", dollar(saved_revenue_1 - campaign_cost_1, prefix = "£")))
cat(sprintf("  • ROI: %.1fx\n\n", saved_revenue_1 / campaign_cost_1))

# Scenario 2: Save 50% of high-risk customers
save_rate_2 <- 0.50
saved_revenue_2 <- revenue_at_risk * save_rate_2
campaign_cost_2 <- total_high_risk * 35  # £35 per customer (more intensive)

cat("Scenario 2: Aggressive (50% save rate)\n")
cat(sprintf("  • Customers saved: %s\n", comma(round(total_high_risk * save_rate_2))))
cat(sprintf("  • Revenue retained: %s\n", dollar(saved_revenue_2, prefix = "£")))
cat(sprintf("  • Campaign cost (£35/customer): %s\n", dollar(campaign_cost_2, prefix = "£")))
cat(sprintf("  • Net benefit: %s\n", dollar(saved_revenue_2 - campaign_cost_2, prefix = "£")))
cat(sprintf("  • ROI: %.1fx\n\n", saved_revenue_2 / campaign_cost_2))

cat(rep("=", 70), "\n\n")
```

---

# Retention Recommendations
```{r recommendations}
cat("TARGETED RETENTION STRATEGIES BY RISK LEVEL\n")
cat(rep("=", 70), "\n\n")

cat("🔴 CRITICAL RISK (Churn Prob >75%)\n")
critical_count <- sum(model_data$Risk_Category == "Critical Risk")
cat(sprintf("   • %s customers\n", comma(critical_count)))
cat("   Actions:\n")
cat("   - Immediate personal outreach (phone calls)\n")
cat("   - Maximum discount offer (30-40% off)\n")
cat("   - Free shipping for 6 months\n")
cat("   - Personal account manager\n")
cat("   - Win-back guarantee (refund if not satisfied)\n")
cat("   Budget: £50-75 per customer\n\n")

cat("🟠 HIGH RISK (Churn Prob 50-75%)\n")
high_count <- sum(model_data$Risk_Category == "High Risk")
cat(sprintf("   • %s customers\n", comma(high_count)))
cat("   Actions:\n")
cat("   - Personalized email campaign\n")
cat("   - 20-25% discount offer\n")
cat("   - Product recommendations based on past purchases\n")
cat("   - Free shipping on next order\n")
cat("   - Feedback survey (understand issues)\n")
cat("   Budget: £25-35 per customer\n\n")

cat("🟡 MEDIUM RISK (Churn Prob 25-50%)\n")
medium_count <- sum(model_data$Risk_Category == "Medium Risk")
cat(sprintf("   • %s customers\n", comma(medium_count)))
cat("   Actions:\n")
cat("   - Automated re-engagement emails\n")
cat("   - 10-15% discount offer\n")
cat("   - \"We miss you\" messaging\n")
cat("   - Highlight new products\n")
cat("   Budget: £10-15 per customer\n\n")

cat("🟢 LOW RISK (Churn Prob <25%)\n")
low_count <- sum(model_data$Risk_Category == "Low Risk")
cat(sprintf("   • %s customers\n", comma(low_count)))
cat("   Actions:\n")
cat("   - Standard marketing automation\n")
cat("   - Loyalty program benefits\n")
cat("   - New product announcements\n")
cat("   Budget: £3-5 per customer\n\n")

cat(rep("=", 70), "\n\n")
```

---

# Export Predictions
```{r export}
# Export churn predictions for CRM
churn_export <- model_data %>%
  select(
    CustomerID,
    Churn_Probability,
    Risk_Category,
    Churned,
    Hist_RecencyDays,
    Hist_TotalTransactions,
    Hist_TotalSpent,
    Hist_AOV,
    Months_Active
  ) %>%
  arrange(desc(Churn_Probability))

export_path <- here("data", "processed", "churn_predictions.csv")
write_csv(churn_export, export_path)

cat("✓ Churn predictions exported to:", export_path, "\n")
cat("  Customers:", nrow(churn_export), "\n\n")

cat("Ready for:\n")
cat("  • CRM integration\n")
cat("  • Retention campaign targeting\n")
cat("  • Customer service prioritization\n")
cat("  • Sales team alerts\n")
```

---

# Summary & Key Takeaways
```{r summary}
cat("\n")
cat(rep("=", 70), "\n")
cat("            CHURN PREDICTION MODEL COMPLETE\n")
cat(rep("=", 70), "\n\n")

cat("✅ ACHIEVEMENTS:\n\n")
cat("  1. Built logistic regression churn prediction model\n")
cat(sprintf("  2. Model accuracy: %.1f%%\n", conf_matrix$overall['Accuracy'] * 100))
cat(sprintf("  3. AUC score: %.3f (%s)\n", auc_value, 
            ifelse(auc_value > 0.8, "Excellent", ifelse(auc_value > 0.7, "Good", "Fair"))))
cat(sprintf("  4. Identified %s high-risk customers\n", comma(total_high_risk)))
cat(sprintf("  5. Quantified %s in at-risk revenue\n\n", dollar(revenue_at_risk, prefix = "£")))

cat("🎯 KEY INSIGHTS:\n\n")
cat(sprintf("  • %.1f%% of customers at high/critical churn risk\n",
            total_high_risk / nrow(model_data) * 100))
cat(sprintf("  • Revenue at risk: %s\n", dollar(revenue_at_risk, prefix = "£")))
cat(sprintf("  • Top risk factor: %s\n", feature_importance$Feature[1]))
cat(sprintf("  • Retention ROI: %.1fx (conservative scenario)\n", saved_revenue_1 / campaign_cost_1))

cat("\n💡 NEXT STEPS:\n\n")
cat("  1. Launch immediate outreach to critical risk customers\n")
cat("  2. Implement risk-based retention campaigns\n")
cat("  3. Integrate churn scores into CRM\n")
cat("  4. Monthly model retraining and updates\n")
cat("  5. Track campaign effectiveness and adjust\n\n")

cat(rep("=", 70), "\n\n")
```

---

# Session Information
```{r session-info}
sessionInfo()
```

---

**✅ Churn Prediction Model Complete!**

*This model enables proactive customer retention and optimized intervention strategies.*